Neural Networks Loss Landscape Convergence in Different Low-Dimensional Spaces

### 2025-04-17 paper/main.tex
Universal commits presents...

\begin{abstract}
  Understanding how the loss landscape of neural networks evolves as the training set size increases is crucial for
  optimizing performance. It is well known that larger datasets can alter the shape of this high-dimensional landscape.
  But the exact point at which additional data no longer brings substantial changes remains underexplored.

  In this paper, we examine neural network models and show that their loss landscapes begin to stabilize once the
  training set grows beyond a certain threshold, revealing a connection between dataset size and the geometry of the
  loss surface. To understand this phenomenon, we propose a method that projects the full parameter space onto a
  low-dimensional subspace derived from top eigenvectors of the Hessian. That provides a more interpretable view of how
  the loss surface in the vicinity of local minima behaves as more data become available. We use different sampling
  strategies, applying Monte-Carlo estimation to capture the structure of this reduced loss landscape more precisely.

  We validate our insights through experiments on image classification tasks, demonstrating that this low-dimensional
  analysis can reveal when the landscape effectively settles, and thus helps determine a minimum viable dataset size.
  Our findings shed light on the relationship between dataset scale and optimization geometry, and suggest
  practical strategies for balancing computational costs with the benefits of additional training data.
\end{abstract}

@@

Neural networks have become irreplaceable in various aspects of life and have a lot of applications now
(e.g. image classification, language models, recommender systems, etc.). However, as datasets and models grow larger and
larger, we get higher accuracy and breakthrough results. However some problems connected with computation resources have
appeared. Sometimes models have so many parameters so we need time we cannot provide for their training. Some attempts to
explore and compare optimization methods have already been made \cite{soydaner2020comparison}. The core problem of all these
solutions hides in their locality, as the size of the neural network and training dataset remains unchanged. In this paper we
explore when adding more data no longer brings significant improvements in network performance and to explain
obtained results in terms of the curvature and shape of the neural network’s loss landscape.

Our research on the loss landscape of neural networks. Specifically, how it changes while adding new samples
to the training dataset. There remains a problem: processing of large datasets requires a lot of time.
We need to determine when increasing dataset size stops reshaping this loss landscape for general neural networks.
But the question of a minimum viable dataset size --- a threshold beyond which further data bring insignificant changes ---
remains underexplored. Moreover, connecting such a threshold to generalization capabilities \cite{wu2017towards} will
reduce computation requirements as we show that there is an upper bound of "active" dataset samples.

We suggest using the Hessian of the loss function to calculate its close approximation. However, the computation of all
Hessian requires a bunch of time, so we consider a projection of neural network parameters space to a certain subspace.
We place the following tasks to the front:
\begin{enumerate}
  \item Constructing a Hessian-based projection of the loss landscape which finds main curvature directions
        (top eigenvectors) in parameter space.
  \item Using different sampling methods, identifying at which dataset size the shape of the loss function stabilizes.
  \item We apply visualization methods to characterize the loss landscape next to the local minima.
  \item Providing applicable theoretical criteria to determine when adding more samples has minimal effect.
        We refer to The Matrix Cookbook \cite{petersen2012matrix} as a hint for matrix operations.
\end{enumerate}

The proposed solution is novel as it links Hessian estimation to the idea of a minimal dataset threshold. This enables
advantages such as more cost-effective dataset collection and a clearer understanding of how dataset size interacts with the
geometry of the loss function. And applying recent methods for low-rank Hessian estimation
can speed up and solve the problem of high computation cost of top eigenvalues for very large networks.

We validate our analysis results on well-known image classification tasks like MNIST \cite{deng2012mnist} and Fashion-MNIST
\cite{xiao2017fashion}. By connecting practical results to theoretical bounds, we offer a method for identifying
a computation-efficiency method that balances performance with accuracy gains.

\textbf{Contributions.} Our contributions can be summarized as follows:
\begin{itemize}
  \item We present a Hessian-based approach that uses a projection into a low-dimensional subspace of top eigenvectors to
        find the critical sufficient dataset size.
  \item We demonstrate the validity of our theoretical framework through empirical studies on MNIST and Fashion-MNIST,
        incrementally increasing dataset size until additional data makes an impact to the loss function curvature.
  \item We highlight the implications of our findings for practical data collection strategies, showing how the detection of
        a sufficient dataset size can reduce computation time.
\end{itemize}

\textbf{Outline.} The rest of the paper is organized as follows. Section 2 divides existing results into some
topics, highlighting their key contributions and findings. Section 3 considers general notation and
some preliminary calculations. In Section 4, we provide theoretical bounds on the hessian and losses
difference norms. Empirical study of the obtained results is given in Section 5. We summarize and
present the results in Sections 6 and 7. Additional experiments and proofs of theorems are included
in the Appendix A.

@@

\textbf{Hessian-based analysis.}
The landscape of loss functions has been explored from various perspectives in the literature. Methods connected with Hessian
are central to understanding the convergence process and loss function landscape \cite{kiselev2024unraveling}. Recent works
observed that the Hessian of some models tends to have a low effective rank, with only a small quantity of eigenvalue
differences from zero. Empirical explorations from \cite{sagun2018empirical} show that the Hessian rank is often very low in
the vicinity of local minima, highlighting a significantly smaller “active subspace”.
Prior work investigating how large eigenvalues emerge during training \cite{ghorbani2019investigation}.

Some results have already been reached for fully connected and convolutional architectures \cite{meshkov2024convnets},
but they need generalization.

\textbf{Loss Landscape Geometry.}
Other recent researches aimed at how dataset size affects both performance and the geometry of loss function landscapes.
In the assumption of enough training samples, neural networks often reach smoother and flatter minima with better
generalization \cite{wu2017towards}. However, collecting large datasets is expensive, as well as the computation
requirements for them, raising questions about the most effective ratio between dataset and model sizes
\cite{hoffmann2022training}. As understanding the landscape of loss functions is key to the solution, prior work has looked
at visualizing loss surfaces \cite{li2018visualizing}.

@@

In this section, we introduce the general notation used in the rest of the paper and the basic assumptions.

We consider a $K$-label classification problem.
So, let's consider $p(y|\bx)$ a conditional probability, which maps unobserved variable $\bx \in \bX$ to the corresponding
output $y \in \by$. We consider $\by$ is a subspace (or same space) as $\mathbb{R}^K$. Let $f_{\bw}(\cdot)$
be a neural network with a list of parameters $\bw$. Let $\Omega$ be a space of parameters ($\bw \in \Omega$).

Let
$$\mathcal{D} = \left\{ (\bx_i, y_i) \, | \, i = 1, \dots, m \right\}$$
be a given dataset of size $m$ consists of i.i.d.
samples. Let $\ell(\cdot, \cdot)$ be a given loss function (e.g. cross-entrophy) where first argument refers to neural network's
result and the second argument refers to the true answer. To simplify, define:
$$\ell_i(\bw) := \ell(f_{\bw}(\bx_i), y_i).$$

\begin{definition}
  The empirical loss function for the first $k$ elements is:
  $$\mathcal{L}_k(\bw) = \frac1M \sum\limits_1^k \ell_i(\bw), \,\, \mathcal{L}(\bw) := \mathcal{L}_m(\bw).$$
\end{definition}

Thus, the difference between losses for neighbouring samplem sizes is:
$$\mathcal{L}_{k}(\bw) - \mathcal{L}_{k-1}(\bw) = \frac{\ell_{k}(\bw) - \mathcal{L}_{k-1}(\bw)}{k}.$$

\begin{definition}
  Let the Hessian of $\mathcal{L}_k(\bw)$ be:
  $$\bH_k(\bw) = \nabla^2_{\bw}\mathcal{L}_k(\bw) = \frac1k \sum\limits_1^k \nabla^2_{\bw} \ell_i(\bw).$$
\end{definition}

\begin{definition}
  To calculate the overall loss landscape changing, one has to integrate the absolute difference for the entire parameter space.
  We define $\mathbf{\Delta}$-\textbf{function} (delta-function) as:
  $$\Delta_{k} = \int \left( \mathcal{L}_{k}(\bw) - \mathcal{L}_{k-1}(\bw) \right)^2 p(\bw) dw,$$
  where $p(\bw)$ describes the priority of the particular parameter points so we can make $p(\bw)$ have higher values
  next to the local minima.
\end{definition}

We firther investigate this difference and aimed at exploration of how adding a new object to the dataset changes the value.
We interested in convergence of this value and properties of loss function when the training dataset size limits to $\infty$.

\begin{definition}
  Let $\Delta$ be a positive hyperparameter that indicates the stop-difference for $\Delta_k$. If
  $$k^* = \inf\limits_k\{\forall m \geq k: \, \Delta_m < \Delta\}$$
  we can say that $k^*$ samples in the dataset are enough to describe the distribution of data from the general population.
  We call $k^*$ as \textbf{sufficient}.
\end{definition}

@@

\begin{assumption}
  Let $\bw^*$ be the local minimum of both $\mathcal{L}_{k-1}(\bw)$ and $\mathcal{L}_{k}(\bw)$.
  Thus, $$\nabla \mathcal{L}_{k-1}(\bw^*) = \nabla \mathcal{L}_{k}(\bw^*) = 0.$$
\end{assumption}

This assumption allows us to explore the behavior and the geometry of the loss function landscape at only one point.

Furthermore, using second-order Taylor's approximation for $\mathcal{L}_{k}(\bw)$ at $\bw^*$ we get:
$$\mathcal{L}_{k}(\bw) \approx \mathcal{L}_{k}(\bw^*) + \frac12 (\bw - \bw^*)^{\T} \bH_k(\bw^*)
  (\bw - \bw^*)$$

\begin{assumption}
  We can assume parameters $\bw$ to be random, which will lead to quite natural condition: $p(\bw)$ can be even a prior
  distribution of $\bw$, so:
  $$\Delta_{k} = \mathbb{E}_{p(\bw)} \left( \mathcal{L}_{k}(\bw) - \mathcal{L}_{k-1}(\bw)
    \right)^2 = \mathbb{D} \left( \mathcal{L}_{k}(\bw) - \mathcal{L}_{k-1}(\bw) \right) + \left( \mathbb{E} \left(
      \mathcal{L}_{k}(\bw) - \mathcal{L}_{k-1}(\bw) \right) \right)^2$$
\end{assumption}

\section{Method}\label{sec:method}

As the work with the full Hessian is computation complex, let's define a projection of neural network parameters space
$\Omega$ to a certain subspace $\Theta$. Let $\bP$ be a projection matrix from $\Omega$ to $\Theta$. Let $\btheta$ be
a projection: $\btheta = \bP\bw$.

Let $\mathbf{V} = \{\mathbf{e}_1, \dots, \mathbf{e}_d\}$ be eigenvectors of $\mathbf{H}$. We select $d$ of them and project parameters space
into a subspace on these $d$ vectors. These vectors should represent maximum eigenvalues of $\mathbf{H}$ as these directions
have the most impact to the loss function difference.
\section{Experiments}\label{sec:exp}

To verify the theoretical estimates obtained, we conducted a detailed empirical study.
In this section, we present experiments aimed at analyzing how the loss landscape evolves as the size of the training
set increases.

\paragraph{Data.}
We use the MNIST dataset, which consists of $60{,}000$ training images and $10{,}000$ test
images of handwritten digits (0--9). Each image is $28\times28$ pixels, grayscale.

\paragraph{Model Architecture.}
We use the following neural network models and preparations:
\begin{itemize}
  \item A multilayer perceptron (MLP) with different number of hidden layers (ReLU activations).
  \item The choice of architecture remains consistent across all experiments.
\end{itemize}

The first two are preliminary experiments, conducted as illustrative examples on the MNIST dataset,
while the others focusing on Hessian-based projections and show connections between eigenvalues and loss function convergence.
That is our primary aim --- to check our theoretical estimates in-deal.

\subsection{Visualizing the Loss Landscape in a Random Two-Dimensional Subspace}
The idea of this initial experiment is to get an intuitive sense of how the loss function’s landscape shifts with
different training-set sizes. To do this, we pick two random directions $\mathbf{v_1}$ and $\mathbf{v_2}$ in the parameter
space and project the network’s parameters onto the subspace they span.

\paragraph{Procedure}
\begin{enumerate}
  \item \textbf{Training on different sizes.} Train a neural network on increasingly larger subsets.
  \item \textbf{Parameter snapshot.} After training, fix the final (or near-final) weights $\mathbf{w}^*$.
  \item \textbf{Subspace grid.} Form a grid of points $\{\mathbf{w}^* + \alpha \mathbf{v}_1 + \beta \mathbf{v}_2\}$ around $\mathbf{w}^*$, where $\alpha$ and $\beta$ vary in a chosen range.
  \item \textbf{Loss evaluation.} Compute and plot the loss at each grid point, producing a 3D visualization.
\end{enumerate}

For smaller $N$, the landscape exhibits sharper fluctuations. As $N$ grows, the landscape often appears smoother and more stable.

\begin{figure}[!htbp]
  \hspace*{-11.5cm}
  \subfloat{\includegraphics[width=2\textwidth]{img/Ls_16.pdf}}
  \caption{Loss function landscape}
  \label{fig:base_loss}
\end{figure}

\begin{figure}[!htbp]
  \hspace*{-11.5cm}
  \subfloat{\includegraphics[width=2\textwidth]{img/Ld_10.pdf}}
  \caption{Loss function difference $(\mathcal{L}_{k} - \mathcal{L}_{k-1})^2$ landscape}
  \label{fig:base_loss_diff}
\end{figure}

\subsection{Visualizing delta-function}

Now we dig deeper and visualize our $\Delta_k$ function to clearly understand the difference in the loss function when
moving from a model trained on $k-1$ samples to one trained on $k$ samples. To do this we sample a bunch of random gaussian
vectors (all components are normally distributed) and calculate a
math expectation of modulus of the loss function difference.
Directly computing this expectation over the entire distribution of random directions $\mathbf{d}$
is impossible, so we employ a Monte Carlo approximation: using Monte-Carlo method (we take the mean value of loss function
on sampled gaussian shifted points):

$$\Delta_k = \mathbb{E}_{p(\bw)} \left( \mathcal{L}_{k}(\bw) - \mathcal{L}_{k-1}(\bw) \right)^2 \approx
  \frac{1}{k} \sum_{t=1}^{k} \left( \mathcal{L}_k(\mathbf{w}^* + \mathbf{d}_t) - \mathcal{L}_{k-1}(\mathbf{w}^* + \mathbf{d}_t) \right)^2$$

where $\mathbf{d}_t \sim \mathcal{N}(0,I_d)$ are $K$ random gaussian samples drawn from a multivariate normal distribution
with zero mean and identity covariance from a subspace of rank $d$. Specifically, we proceed as follows:

\begin{enumerate}
  \item \textbf{Sampling Directions.} Generate $K$ independent directions $\mathbf{d}_1, \dots, \mathbf{d}_M$
        from $\mathcal{N}(0,I_d)$.
  \item \textbf{Decreasing multidimension.} Project the network’s parameters onto the subspace they span.
  \item \textbf{Loss Evaluation.} For each direction $\mathbf{d}_t$, evaluate the loss at
        $\mathbf{w}^* + \mathbf{d}_t$ and
        $$\left( \mathcal{L}_{k}(\bw + \mathbf{d}_t) - \mathcal{L}_{k-1}(\bw + \mathbf{d}_t) \right)^2 $$
  \item \textbf{Averaging.} Average these values values over $k = 1, \dots, K$ to approximate $\Delta_k$.
\end{enumerate}

By increasing $K$, we obtain a more reliable estimate of the expected difference in loss landscapes.
This Monte Carlo approach is straightforward to implement and does not require explicit knowledge of the underlying loss
function's distribution.

The $\Delta$-function generally decreases as $N$ increases, suggesting that the landscape differences between neighbouring
sizes become smaller.
Beyond a certain point, $\Delta$-values are nearly negligible, indicating that adding more data has minimal impact on
the local landscape in these directions.

\noindent
\textbf{Conclusion from Preliminary Experiments.}
These initial experiments demonstrate that the loss landscape becomes progressively more stable with increasing $N$.
However, the directions used here are either random. This motivates a more rigorous approach in the \emph{main experiment}
below, where we focus specifically on the most critical directions of curvature --- those corresponding to the top
eigenvalues of the Hessian.

\subsection{Main Experiment: Projection onto Dominant Hessian Eigenvectors}

From prior work \cite{sagun2018empirical}, we know that only a few Hessian eigenvectors (those associated with the largest
eigenvalues) tend to capture the most significant curvature directions. By projecting onto these principal directions,
we can more precisely measure how the landscape the loss surface changes as $k$ increases.

\paragraph{Procedure}
\begin{enumerate}
  \item \textbf{Hessian approximation.} For each trained model, approximate
        the Hessian $\mathbf{H}$ or its dominant eigenvalues/eigenvectors. Techniques ???
  \item \textbf{Eigen-decomposition.} Identify the top $d$ eigenvectors $\{\mathbf{e}_1, \dots, \mathbf{e}_d\}$ with the
        largest eigenvalues.
  \item \textbf{Calculation and analization.} For each model, calculate a $\Delta$ fuction via the same way as in the second
        experiment (using Monte-Carlo) and visualize via plots.
\end{enumerate}

\paragraph{Goal and Expected Outcome}
\begin{itemize}
  \item By focusing on the directions of largest curvature, we expect a clearer
        measure of how the landscape of the loss surface changes with additional data.
  \item This approach should yield a more accurate estimate of the point at which the
        landscape effectively stabilizes, thus helping to pinpoint a minimum viable dataset size.
  \item By applying experiments we check our calculations above on the reality.
\end{itemize}

\begin{figure}[!htbp]
  \centering
  \subfloat{\includegraphics[width=1\textwidth]{img/D_32.pdf}}\\
  \caption{$\Delta$-function visualization. `Sigma` parameter refers to the variance of sampled gaussian vectors, `dim'
    refers to the rank of the subspace.}
  \label{fig:base_delta}
\end{figure}

>>>>

\begin{abstract}
  Understanding how a neural network’s loss landscape changes as we add more training data is important for efficient training.
  Although larger datasets are known to reshape this high-dimensional surface, the point when extra data stop making a big difference
  is still unclear.

  In this paper, we study this issue and show that the loss landscape near a local minimum stabilizes once the dataset exceeds a
  certain size. To analyze this, we project the full parameter space onto a smaller subspace formed by the Hessian’s top eigenvectors.
  This low-dimensional view highlights how the loss surface changes in its most important directions. We then apply Monte Carlo sampling
  within this subspace to estimate these changes more precisely.

  We test our approach on standard image‑classification tasks and find that our low-dimensional analysis pinpoints when the landscape
  stops evolving. These findings clarify how dataset size affects optimization and offer practical guidance for balancing training cost
  with performance gains.
\end{abstract}

@@

Neural networks are now essential in many areas --- image classification, language modeling, recommender systems, and more. As models and
datasets grow, we often see better accuracy and new breakthroughs. But bigger networks and data also demand more computation time and
resources, which can become prohibitive. While prior work has compared various optimization methods under fixed data and model sizes
\cite{soydaner2020comparison}, the question of when adding more training samples stops yielding significant gains remains largely
unanswered.

In this paper, we investigate exactly that: how the loss landscape changes as we increase the dataset size, and at what point further
data have little effect. Knowing this “minimum viable dataset size” --- a threshold beyond which new samples bring negligible
improvement --- can save both training time and data‑collection effort. We also relate this threshold to generalization behavior, along
the lines of \cite{wu2017towards}.

To study this, we use the Hessian of the loss function as a proxy for local curvature. Computing the full Hessian is expensive, so we
project the parameter space onto a low‑dimensional subspace formed by its top eigenvectors. Our main contributions are:

\begin{enumerate}
  \item Constructing a Hessian‑based projection that captures the principal curvature directions.
  \item Applying Monte Carlo and other sampling methods to identify when the loss landscape stabilizes as dataset size grows.
  \item Visualizing the projected loss surface around local minima to illustrate these changes.
  \item Deriving theoretical criteria --- using results from The Matrix Cookbook \cite{petersen2012matrix} --- to predict when additional
        samples have minimal impact.
\end{enumerate}

This approach is novel in linking low‑rank Hessian estimation to a concrete dataset threshold, enabling more cost‑effective data
collection and a clearer understanding of how data scale interacts with loss geometry. We validate our method on MNIST
\cite{deng2012mnist} and Fashion‑MNIST \cite{xiao2017fashion}, demonstrating how to balance computational cost against accuracy gains.

\textbf{Organization.} The rest of the paper is structured as follows. Section~2 reviews related work. Section~3 introduces notation
and preliminaries. Section~4 derives theoretical bounds on Hessian spectra and the loss‐difference norm. Section~5 presents our
empirical studies. We discuss the implications in Section~6 and conclude in Section~7. Additional experiments and proofs are
provided in Appendix~A.

@@

\textbf{Hessian-based Analysis.}
The loss landscape of neural networks has been extensively studied using Hessian-based techniques, which illuminate convergence
behavior and curvature properties \cite{kiselev2024unraveling}. Empirical work shows that near local minima the Hessian often has low
effective rank, indicating a small "active subspace" spanned by a few large eigenvalues \cite{sagun2018empirical}. Ghorbani et al.
\cite{ghorbani2019investigation} examined how these large eigenvalues emerge during training and shape optimization dynamics.
Meshkov et al. \cite{meshkov2024convnets} extended this analysis to fully connected and convolutional architectures, though broader
validation across architectures remains an open challenge.

\textbf{Loss Landscape Geometry and Dataset Size.}
Another research direction explores how dataset size influences model performance and landscape geometry. Networks trained on larger
datasets tend to find flatter, wider minima that generalize better \cite{wu2017towards}. Yet, acquiring and processing massive data is
costly, motivating studies of optimal data–model trade‑offs \cite{hoffmann2022training}. Li et al. \cite{li2018visualizing} developed
visualization tools that reveal how dataset size, initialization, and architecture affect the topology of loss surfaces.

\textbf{Algorithmic Stability and Landscape Sensitivity.}
Algorithmic stability, which measures how small changes in the training set affect the learned model, provides theoretical bounds on
generalization error \cite{bousquet2002stability}. Elisseeff et al. \cite{elisseeff2005stability} extended stability analysis to
randomized algorithms such as bagging and ensemble methods, deriving non-asymptotic generalization guarantees. In our context,
stability theory underpins the intuition that, beyond a certain dataset size, further samples cause negligible changes to the loss
curvature. Our Monte Carlo–based Hessian projection method directly leverages this perspective, using top eigenvectors to quantify
and visualize when the loss landscape stabilizes as new data are added.

@@

We consider a $K$-class classification problem. Let $\bx \in \bX$ be an input and $y \in \mathcal Y = \{1,\dots,K\}$ its label. A
neural network with parameters $\bw \in \bOmega \subset \mathbb R^P$ defines a mapping $f_{\bw} : \bX \to \mathbb R^K$. Given a dataset
$$
  \mathcal D =
  \{(\bx_i, y_i)\}_{i=1}^m
$$
of $m$ i.i.d. samples, and a per-sample loss $\ell \bigl( f_{\bw}(\bx_i), y_i \bigr)$ (e.g. cross–entropy), we define
$$
  \ell_i(\bw) =
  \ell \bigl( f_{\bw}(\bx_i) , y_i \bigr)\,.
$$

\begin{definition}
  The empirical loss on the first $k$ samples is
  $$
    \cL_k(\bw) =
    \frac1k \sum_{i=1}^k \ell_i(\bw),
  $$
  so that the full-sample loss is $\cL(\bw) = \cL_m(\bw)$.
\end{definition}

A simple telescoping gives the difference
$$
  \cL_{k+1}(\bw) - \cL_k(\bw) =
  \frac{\ell_{k+1}(\bw) - \cL_k(\bw)}{k+1}\,.
$$

\begin{definition}
  The Hessian of $\cL_k$ at $\bw$ is
  $$
    \bH_k(\bw) =
    \nabla^2_{\bw} \cL_k(\bw) =
    \frac1k \sum_{i=1}^k \nabla^2_{\bw} \ell_i(\bw),
  $$
  and, similarly, the full-sample Hessian is $\bH(\bw) = \bH_m(\bw) = \nabla^2_{\bw} \cL_m(\bw)$.
\end{definition}

To capture the overall change in the landscape when adding one more sample, we introduce a $\Delta$‑function:

\begin{definition}
  Let $p(\bw)$ be a weighting (e.g. peaked near a local minimum). Define
  $$
    \Delta_k =
    \int \bigl( \cL_{k+1}(\bw) - \cL_k(\bw) \bigr)^2\, p(\bw)\, d\bw.
  $$
\end{definition}

We study the behavior of $\Delta_k$ as $k \to \infty$ and define a threshold:

\begin{definition}
  Fix a tolerance $\Delta > 0$. The \emph{sufficient sample size} is
  $$
    k^* =
    \inf \{ k \mid \forall\, m \ge k : \; \Delta_m < \Delta\}.
  $$
  When $k \ge k^*$, adding further samples changes the loss landscape by less than $\Delta$.
\end{definition}

@@

We make the following standard assumptions to simplify our analysis around a local minimizer~$\bw^*$.

\begin{assumption}
  The point $\bw^*$ is a common local minimizer for both $\cL_k$ and $\cL_{k+1}$, i.e.
  $$
    \nabla \cL_k(\bw^*) =
    \nabla \cL_{k+1}(\bw^*) =
    0.
  $$
\end{assumption}

Under this assumption, a second‑order Taylor expansion of $\cL_k$ about $\bw^*$ gives
$$
  \cL_k(\bw) \approx
  \cL_k(\bw^*) + \tfrac12\, (\bw - \bw^*)^{\T} \bH_k(\bw^*)\, (\bw - \bw^*).
$$

\begin{assumption}
  At the minimizer $\bw^*$, the loss values for consecutive sample sizes are the same:
  $$
    \cL_{k+1}(\bw^*) =
    \cL_k(\bw^*).
  $$
\end{assumption}

\begin{assumption}
  The parameters $\bw$ are distributed according to a density $p(\bw)$ (e.g. a Gaussian prior centered at $\bw^*$).  Then
  $$
    \Delta_k =
    \int \bigl(\cL_{k+1}(\bw)-\cL_k(\bw)\bigr)^2\, p(\bw)\, d\bw =
    \mathbb{E}_{p(\bw)} \left( \cL_{k+1}(\bw) - \cL_{k}(\bw) \right)^2 =
  $$
  $$
    \mathbb D_{p(\bw)} \bigl[ \cL_{k+1}(\bw) - \cL_k(\bw) \bigr] +
    \Bigl( \mathbb E_{p(\bw)} \bigl[ \cL_{k+1}(\bw) - \cL_k(\bw) \bigr] \Bigr)^2.
  $$
\end{assumption}

\section{Method: Projection onto Dominant Eigen-Directions and Loss Landscape Approximation}

Working with the full Hessian $\mathbf{H}_k(\bw)$ of a neural network is computationally expensive due to the high dimensionality
of the parameter space $\bOmega$. To reduce this cost, we project the parameters onto a lower‑dimensional subspace spanned by the
$d$ dominant eigenvectors of the Hessian at a local minimum $\bw^*$.

The justification for this projection is that, empirically, the Hessian has low effective rank near a minima \cite{sagun2018empirical}:
only a few eigenvalues are significantly above zero, while the rest are negligible.  Therefore, directions corresponding to small
eigenvalues contribute little to loss variation.  Retaining only the top $d$ eigenvectors yields a compact yet informative representation
of the most important curvature directions.

For illustration, Figure~\ref{fig:evgen} plots the top few eigenvalues of $\bH(\bw^*)$ in descending order. We compute these dominant
eigenvalues using the iterative power‐method described at \cite{hessian-eigenthings}.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.7\textwidth]{img/eigenvalues.pdf}
  \caption{Top eigenvalues of $\bH(\bw^*)$ in descending order.}
  \label{fig:evgen}
\end{figure}

Concretely, let
$$
  \bP =
  \bigl[ \mathbf{e}_1, \dots, \mathbf{e}_d \bigr]
$$
be the matrix whose columns are the top $d$ eigenvectors.  We write any parameter vector as
$$
  \bw =
  \bw^* + \bP\, \btheta, \quad
  \btheta \in \mathbb R^d.
$$
Substituting into a second‑order Taylor expansion of $\cL_k$ around $\bw^*$ gives
$$
  \cL_k\bigl(\bw^* + \bP\,\btheta\bigr) \approx
  \cL_k(\bw^*) + \tfrac12\, \btheta^{\T}\! \bigl( \bP^{\T}\, \bH_k(\bw^*)\, \bP\bigr)\, \btheta =
  \cL_k(\bw^*) + \tfrac12\, \btheta^{\T} \bLambda_k\, \btheta,
$$
where
$$
  \bLambda_k =
  \bP^{\T}\, \bH_k(\bw^*)\, \bP =
  \operatorname{diag} \bigl( \lambda_k^1, \dots, \lambda_k^d \bigr).
$$
An identical expansion holds for $\cL_{k+1}$, yielding $\bLambda_{k+1}$ from the Hessian on $k+1$ samples.

\begin{theorem}[Approximate $\Delta_k$ via Eigenvalues]
  Under the assumption that $\btheta \sim \mathcal N(0, \sigma^2 \mathbf I_d)$, the change in loss can be estimated by
  $$
    \Delta_k =
    \mathbb E \bigl[ \cL_{k+1}(\bw) - \cL_k(\bw) \bigr]^2 \approx
    \frac{\sigma^4}{4}\, \Biggl( 2 \sum_{i=1}^d (\lambda_{k+1}^i - \lambda_k^i)^2
    + \Bigl( \sum_{i=1}^d (\lambda_{k+1}^i - \lambda_k^i) \Bigr)^2 \Biggr).
  $$
\end{theorem}

\begin{remark}
  Assuming $\mathbb{E}_{p(\btheta)}(\btheta) = \mathbf{0}$ is natural when centering the sampling distribution at the minimum.
\end{remark}

This formula uses only the top $d$ eigenvalues, so it is efficient to compute even for very large models.
In practice, it provides an upper bound on the empirical $\Delta_k$ and thus a reliable criterion for determining when the
dataset size is sufficient.

\section{Experiments}

To validate our theoretical estimates, we perform a set of experiments that track how the loss landscape changes as we grow the
training set.

\paragraph{Datasets.}
We evaluate on MNIST \cite{deng2012mnist} and Fashion‑MNIST \cite{xiao2017fashion}, each containing 60,000 training and 10,000 test
grayscale images of size $28\times28$ pixels.

\paragraph{Model Architecture.}
All experiments use a simple multilayer perceptron (MLP) with ReLU activations.  We keep the same architecture—number of hidden layers
and hidden units—across all runs to ensure a fair comparison.

\paragraph{Experimental Procedure.}
\begin{enumerate}
  \item \textbf{Random‐direction visualization:} As an initial sanity check, we project the parameters onto two random directions and
        plot the resulting 3D loss landscapes and their squared differences.
  \item \textbf{Hessian‐based analysis:} We then repeat the visualization and compute $\Delta_k$ by projecting onto the top two
        (and up to $d = 10$) eigenvectors of the Hessian.  This directly tests our eigenvalue‐based approximation and the convergence behavior
        predicted by theory.
\end{enumerate}

All code and detailed configuration files are available in our GitHub repository.

\subsection{Preliminary Experiment: Projection onto Random Directions}
\subsubsection{Visualizing the Loss Landscape}

To build intuition for how the loss surface changes with training‐set size, we first project the model parameters onto two random
directions. Specifically, we pick two random vectors $\mathbf v_1, \mathbf v_2 \in \mathbb R^P$ in parameter space and study the
subspace they span.

We then evaluate the empirical loss
$$
  \cL_k(\bw) =
  \frac1k \sum_{i=1}^k \ell_i(\bw),
$$
where $\ell_i(\bw) = \ell (f_\bw(\bx_i), y_i)$, over a grid in the $(\alpha, \beta)$ plane defined by
$$
  \bw(\alpha,\beta) =
  \bw^* + \alpha\, \mathbf v_1 + \beta\, \mathbf v_2, \quad
  (\alpha, \beta) \in [-1,1]^2.
$$

\begin{enumerate}
  \item \textbf{Grid Sampling.}
        We form a uniform grid of $(\alpha, \beta)$ pairs in $[-1,1]\times[-1,1]$ with a fixed step size.
  \item \textbf{Loss Computation.}
        For each $(\alpha,\beta)$, we set $\bw=\bw(\alpha,\beta)$, keep the base weights $\bw^*$ fixed, and compute both
        $\cL_k(\bw)$ and $\cL_{k+1}(\bw)$.
  \item \textbf{Aggregated Metrics.}
        \begin{itemize}
          \item The surface $\cL_k(\alpha,\beta)$ visualizes how the mean loss changes over the grid.
          \item The surface $\bigl( \cL_{k+1} (\alpha, \beta) - \cL_k (\alpha, \beta) \bigr)^2$ highlights where adding one sample
                causes the largest change.
        \end{itemize}
  \item \textbf{3D Visualization.}
        We render two side‐by‐side 3D plots:
        \begin{itemize}
          \item \emph{Left:} $\cL_k (\alpha, \beta)$ vs. $(\alpha, \beta)$.
          \item \emph{Right:} $\bigl( \cL_{k+1} (\alpha, \beta) - \cL_k (\alpha, \beta) \bigr)^2$ vs. $(\alpha, \beta)$.
        \end{itemize}
\end{enumerate}

\begin{figure}[!htbp]
  \hspace*{-2.2cm}
  \subfloat{\includegraphics[width=1.3\textwidth]{img/loss_random_1_2.pdf}}
  \caption{Loss function landscape for small $k$, projected onto the two random directions.}
  \label{fig:loss_random_small}
\end{figure}

\begin{figure}[!htbp]
  \hspace*{-2.2cm}
  \subfloat{\includegraphics[width=1.3\textwidth]{img/loss_random_-2_-1.pdf}}
  \caption{Loss function landscape for maximum $k$, projected onto the two random directions.}
  \label{fig:loss_random_big}
\end{figure}

\paragraph{Loss Landscape for Small Sample Size}
In Figure~\ref{fig:loss_random_small} ($k=1$), the loss surface is highly irregular, with sharp peaks and deep valleys.
This irregularity indicates that with very few training samples, the model is extremely sensitive to small parameter perturbations.
The squared difference surface also exhibits large spikes, confirming that adding a single data point can cause substantial local changes
in the loss.

\paragraph{Loss Landscape for Maximum Dataset Size}
In contrast, Figure~\ref{fig:loss_random_big} (maximum $k$) shows a much smoother loss surface with gentle slopes. The overall loss
values are lower --- reflecting the averaging effect of more samples --- and the squared difference surface is correspondingly much smaller.
This smoothing demonstrates that as the dataset grows, the local optimum stabilizes, and additional samples produce only minor
adjustments to the loss landscape. Hence, beyond a certain dataset size, further data have negligible impact on the shape of the loss
surface.

\subsubsection{Visualizing the \texorpdfstring{$\Delta_k$}{Delta k} Function}

To study how the loss landscape converges as the dataset grows, we estimate and plot the function
$$
  \Delta_k =
  \mathbb{E}_{p(\bw)} \bigl[ \cL_{k+1}(\bw) - \cL_k(\bw) \bigr]^2
$$
via a Monte Carlo approximation using Gaussian perturbations in parameter space. The steps are:

\begin{enumerate}
  \item \textbf{Sample directions.}
        Starting from a fixed trained model with parameters $\bw^*$, draw $K$ independent random vectors
        $$
          \btheta^{(t)} \sim \mathcal{N} (\mathbf 0, \sigma^2 \mathbf I_d), \quad
          t = 1, \dots, K.
        $$

  \item \textbf{Perturb parameters.}
        For each sample, form a perturbed parameter set
        $$
          \bw^{(t)} =
          \bw^* + \btheta^{(t)}.
        $$

  \item \textbf{Evaluate loss differences.}
        Compute
        $$
          \cL_k \bigl( \bw^{(t)} \bigr) \quad
          \text{and} \quad
          \cL_{k+1} \bigl( \bw^{(t)} \bigr),
        $$
        then calculate the squared difference
        $$
          d_t =
          \bigl[ \cL_{k+1} (\bw^{(t)}) - \cL_k (\bw^{(t)}) \bigr]^2.
        $$

  \item \textbf{Compute Monte Carlo estimate.}
        Approximate
        $$
          \Delta_k \approx
          \frac{1}{K} \sum_{t=1}^K d_t.
        $$
        Increasing $K$ reduces the variance of this estimator.
